Separation of an upwelling current bounding the Juan de Fuca Eddy

### 2022-06-07 PW13Paper.tex
Author
FIX
DOC: version after SNW suggestions

\title{Separation of an upwelling current bounding the Juan de Fuca Eddy, off Vancouver Island}

@@

\authors{Jody M. Klymak\affil{1} and Everyone }

% \affiliation{1}{First Affiliation}
% \affiliation{2}{Second Affiliation}
% \affiliation{3}{Third Affiliation}
% \affiliation{4}{Fourth Affiliation}
\affiliation{1}{School of Earth and Ocean Sciences and Department of Physcics \& Astronomy, University of Victoria, B.C.\ Canada}
%(repeat as many times as is necessary)

@@

\begin{keypoints}
\item The shelfbreak current along Vancouver Island separates downstream of a submarine bank.
\item Offshore water is drawn onto the shelf and forms a sharp semi-persisnte front with the Juan de Fuca eddy.
\item The Juan de Fuca eddy shows evidence of long residence times, and little evidence of deep-water origin.
\end{keypoints}

%% ------------------------------------------------------------------------ %%
%
%  ABSTRACT and PLAIN LANGUAGE SUMMARY
%
% A good Abstract will begin with a short description of the problem
% being addressed, briefly describe the new data or analyses, then
% briefly states the main conclusion(s) and how they are supported and
% uncertainties.

% The Plain Language Summary should be written for a broad audience,
% including journalists and the science-interested public, that will not have
% a background in your field.
%
% A Plain Language Summary is required in GRL, JGR: Planets, JGR: Biogeosciences,
% JGR: Oceans, G-Cubed, Reviews of Geophysics, and JAMES.
% see http://sharingscience.agu.org/creating-plain-language-summary/)
%
%% ------------------------------------------------------------------------ %%

%% \begin{abstract} starts the second page

@@

@@

Cross-shelf exchange is important to the health and productivity of continental shelf regions, where oxygenated water is exchanged with nutrient-rich, but oxygen-depleted nearshore water.  Cross-shelf transport usually requires ageostrophic flow, since balanced flow will tend to follow topographic contours, often providing a barrier to lateral exchange \cite{brink16}.  Mechanisms for cross-shelf exchange include internal waves and instabilities in shelf-break fronts.  Most-dramatically, however, topography can drive cross-slope exchanges, often seen most clearly in surface imagery, but occasionally observed \emph{in-situ} \cite{barthetal00}.

Here we present detailed observations from the southern Vancouver Island shelf, collected in summer 2013, and sampled with a rapid profiling vehicle equipped with a CTD and oxygen sensor, supplemented by traditional hydrographic surveys. The shelf has complicated bathymetry \fref{fig:LocMapBoth}, with a somewhat typical shelf poleward of our study site, but then widening in the lee of La Perouse Bank, and then hitting the Juan de Fuca canyon, equatorward.  Water from the strait of Juan de Fuca flows poleward as a buoyant current that hugs the coast \cite{thomsonetal89, hickeyetal91}, while shelf water flows equatorward in the summer under upwelling conditions, both forced by local winds and via tele-connections with the long, homogenous shelves equatorward off Washington and Oregon \cite{hickeyetal91,thomsonkrassovski15,engidaetal16}.  Trapped between these two currents is a region of relatively homogenous water that has been termed the ``Juan de Fuca'' eddy \cite{freelanddenman82,freelandmcintosh89,foremanetal07,macfadyenhickey10}.

One of our findings below is that the water in the deep pool of the Juan de Fuca eddy is either  quite old or undergone significant vertical mixing.  The low oxygen found in the Juan de Fuca eddy has caused it to be inferred to be upwelled from as deep as 400 m \cite{freelanddenman82,deweycrawford88}.  This inference was based on the oxygen being conservative, and hence having to come from close to the oxygen minimum zone found further offshore \cite{mackasetal87}.  This lead to hypotheses that perhaps the eddy was fed by waters being drawn up a spur canyon of the Juan de Fuca canyon (called the Spur Canyon) via ageostrophic transport from offshore to onshore due to the low pressure in the eddy center \cite{weaverhsieh87}.  Below, we will argue that there is not substantial evidence of such transport, and that oxygen is likely low because of consumption due to respiration.

A second important finding is that the water flowing equatorward along the shelf from upstream all detaches from the shelf and is pushed offshore to be replaced by offshore water that floods the southern shelf.  Such events are important as they tend to transport nutrient rich shelf water out into the open ocean, fueling offshore ecosystems.  Similar offshore motion has been remarked in the region using satellite images \cite{ikedaemery84,thomsongower98}, and observed further to the south off Cape Blanco \cite{barthetal00}, and other coasts \cite[e.g.]{relvasbarton05}. Net cross-shore transport of nutrients and chlorophyll have been found in hydrographic surveys off Vancouver Island \cite{mackasyelland99}, and associated with mesoscale features both in geostrophic velocities, and in satellites.  The mechanisms driving such separations are poorly understood, but hypotheses include coastal hydraulics \cite{dalebarth01} leading to along-shore trapping of coastal waves,  wind stress curl variations \cite{castelaobarth07}, induced relative vorticity due to stretching of parcels that inertially overshoot their initial isobaths due to a sudden change in downstream bathymetry, and the non-linear breaking of large-scale meanders due to baroclinic instability between the wind-driven current and the California Undercurrent \cite{ikedaetal84, batteen97}.

In this paper we present  observations from 2013 (\fref{sec:Site}), presenting the data in a number of ways to highlight the important processes, with a focus on the offshore side of the eddy region (\fref{sec:Observations}) where we consider the age of the eddy, the steadiness of the front separating the eddy and the offshore water, properties along the spur canyon that was believed to feed the eddy, and finally describe the large offshore separation of the shelf-break current and its intrusion into the interior.  We discuss the origin of the eddy and the implications of the separating jet (\fref{sec:Summary}).

@@

Observations from the LB and LC lines were collected aboard the \emph{CCGS Tully} from 2013-05-30 to 2013-05-31, which we will call ``June'', and 2013-09-07 to 2013-09-09, which we will call ``September''.  These lines span the depths from about 50 m to deep offshore.  The data comes from a lowered Seabird 9-11 CTD, with an SBE 43 oxygen sensor.  Oxygen data have been checked against bottle casts, and the CTD corrected for sensor offsets and thermal lags.

Observations aboard the \emph{R/V Falkor} were carried out from 2013-08-21 to 2013-08-30, starting two weeks before the September \emph{Tully} cruise.  Data were collected with an ODIM Brooke-Ocean Moving Vessel Profiler, profiling to 200 m.  The MVP was equipped with an AML Oceanographic CTD, and a Rinko Oxygen sensor with a 7-s response time foil.  The MVP profiles at vertical fall speeds of $3\ \mathrm{m\,s^{-1}}$ to depths of 200 m or to within 5 m of the seafloor, whichever is shallower.  Data collection took place while the ship cruised at speeds between 5 and 8 kts, usually at around 6 kts to enable fine horizontal spacing of the casts, with typical spacing less than 800 m.  Data is collected during down- and up-casts, but is only reported for down-casts because the path through the water is highly asymmetric on the up-cast, whereas the downcast is largely vertical.  The rapid speed of the profiling makes the oxygen measurements somewhat coarse, and probably biased due to the phase lag of the sensor, so we treat these largely qualitatively in this paper.

@@

Winds during the cruises were typical for the west coast of Vancouver Island, with upwelling favorable winds during July and early August (\fref{fig:LaPeWind}).  During the cruise, and for the week previous, the winds were intermittently downwelling favorable.  Note that this locale is strongly affected by coastally trapped waves from further south, so doming of near-bottom isopycnals often persists despite local wind forcing \cite{thomsonkrassovski15, engidaetal16}, and takes a finite amount of time to spin down.

\begin{figure*}[htbp]
  \begin{center}
    \includegraphics[width=5.5in]{LaPeWind}
    \caption{
      Wind from the La Perouse buoy \cite{DFOWind2013C46206}.  The vertical direction is along-shelf (chosen as a heading of 315 N).  The wind components has been low-pass filtered to one-day averages.  The timing of the cruises discussed in the paper are shown as colored bands.
      \label{fig:LaPeWind} }
  \end{center}
\end{figure*}

@@

Hydrographic sections along the LB and LC lines highlight summer conditions on the Southern Vancouver Island Shelf, and indicate some of the features we are focusing on in this paper (\fref{fig:LaPerouse2013Ctd}).  During both cruises and at both lines there is clear evidence of upwelling, with the $26.4\ \mathrm{kg\,m^{-3}}$ isopycnal reaching from 130 m to shallower than 85 m over the shelf.  This dense water tends to be low in oxygen and cool. Further onshore, isopycnals tilt back down, and warm water fills the water column, evidence of the Vancouver Island Coastal Current that hugs the coast here.

\begin{figure*}[htbp]
  \begin{center}
     \includegraphics[width=130mm]{LaPerouse2013Ctd}
     \caption{Observations along LC and LB lines in Jun and Sep 2013.  X is across-shelf distance in the coordinate system shown in \fref{fig:LocMapBoth}. Potential density is contoured every $0.2\ \mathrm{kg\,m^{-3}}$, with the $26.4\ \mathrm{kg\,m^{-3}}$.  Potential temperature $\theta$, spice anomaly $\gamma$ and oxygen concentration $O_2$ are colored for each section.}
     \label{fig:LaPerouse2013Ctd}
  \end{center}
\end{figure*}

The ``Juan de Fuca Eddy'' region is along the LB Line, mostly inshore of 0 km. This water is cooler and lower in oxygen than along the same isopycnal further offshore.  Also note that during the June cruise the water in this region was cooler and more oxygenated than later in the summer during the September cruise.  Near the surface, the isopycnals are domed, but the low-oxygen anomaly extends as high as the thin surface mixed layer.

In contrast, the water upstream, along the LC line, does not show as much drawdown of oxygen, and is somewhat warmer, at least offshore of La Perouse bank (\fref{fig:LaPerouse2013Ctd} top three rows).  Water does not appear to make it over La Perouse bank into the basin onshore, or if it does, it does so intermittently.  There appears to be some mixing of the deeper water and shallow water near $26.4\ \mathrm{kg\,m^{-3}}$, with the water along this isopycnal becoming cooler and lower in oxygen where it intersects the shelf.

The $\theta$--$S$ properties show many of the features demonstrated above (\fref{fig:LaPerouse2013TS}a).  Below $26.6\ \mathrm{kg\,m^{-3}}$ or so, the deep water masses are found offshore, and are largely homogenous. Shallower, there are distinct differences between the water found on the shelf and water found more offshore.  These anomalies are hard to parse directly, so in this paper we use a spice anomaly, relative to a straight line in $\theta$--$S$ space (\fref{fig:LaPerouse2013TS}). The line passes the points $7.75\ \mathrm{^oC}$, $30\ \mathrm{psu}$ and $6.6\ \mathrm{^oC}$, $35\ \mathrm{psu}$, and spice anomaly for a given water sample at density $\sigma_{\theta}$ is $\gamma = \alpha \left(T - T_0\left(\sigma_{\theta}\right)\right) + \beta \left(S - S_0\left(\sigma_{\theta}\right)\right)$ where $T_0\left(\sigma_{\theta}\right)$ and $S_0\left(\sigma_{\theta}\right)$ are the temperature and salinity along the mixing line at the same density as the water sample, with the sign convention that positive is warmer and saltier than data along the mixing line.  It is hard to see from the $\theta$--$S$ plot in \fref{fig:LaPerouse2013TS}, but this mixing line is a strong feature of the water masses discussed in this paper.

\begin{figure}[htbp]
  \begin{center}
     \includegraphics[width=5in]{LaPerouse2013TS}
     \caption{Potential temperature versus salinity for the La Perouse data shown in \fref{fig:LaPerouse2013Ctd}.  Potential density is contoured in the same way.  A definition of spice anomaly is shown in this plot, and discussed in the text.  The rectangle is the $\theta-S$ range used in \fref{fig:TSdensSpice} below.}
     \label{fig:LaPerouse2013TS}
  \end{center}
\end{figure}

The spice anomaly defined this way is mapped in the middle row of each set of plots in \fref{fig:LaPerouse2013Ctd}.  Offshore water tends to be high is absolute spice anomaly, with a positive-anomaly layer (red) centered at $26.4\ \mathrm{kg\,m^{-3}}$ sandwiched between negative-anomaly layers above and below.  On the shelf ($-10\, \mathrm{km} < X < 30\, \mathrm{km}$), the distinct $\theta$--$S$ masses are attenuated and much closer to the mixing line than the offshore water.  Hugging the coast ($X > 30\, \mathrm{km}$), is the Vancouver Island Coastal Current, with a negative spice anomaly in June, and a positive anomaly in summer, consistent with the warming of this water during the summer.

@@

     \includegraphics[width=3.5in]{TSdensSpice.pdf}
    \caption{Sample density of salinity and potential density data from the cruise, with a logarithmic color scale.  Grey contours are potential density relative to the surface, every $0.1\ \mathrm{kg\,m^{-3}}$, and the magenta contour is the $26.4\ \mathrm{kg\,m^{-3}}$ isopycnal.  Spice anomaly, $\gamma$ is defined as temperature anomaly scaled by $\alpha$ along isopycnals from the mixing line found in the well-mixed region onshore, in $\mathrm{kg\,m^{-3}}$.
    \label{fig:TSdensSpice}}
  \end{center}
\end{figure*}

These three water masses are quite coherent in space.  While we did not do the specific La Perouse sections, our spatial coverage is such that we can make consistent synthetic sections (\fref{fig:CrossSectionsSpice}), taken at $y=42, 28, 14$, and $0\, \mathrm{km}$ (see \fref{fig:LocMapBoth}).  During the August cruise, the upstream water along $\sigma_{\theta} = 26.4\ \mathrm{kg\,m^{-3}}$ has moderate spice anomaly of  $\gamma \approx 0.1\,\mathrm{kg\,m^{-3}}$ (\fref{fig:CrossSectionsSpice}a), and in this representation, the water appears "pink".  There is some higher-anomaly water offshore of $x \approx -40\, \mathrm{km}$ along this upstream section.

By $y\approx 14\, \mathrm{km}$, this ``pink'' water along $\sigma_{\theta} = 26.4\,\mathrm{kg\,m^{-3}}$ is gone and is replaced by high spice anomaly water $\gamma \approx 0.2 \,\mathrm{kg\,m^{-3}}$ (``red'') that is directly next to well-mixed water ($\gamma \approx 0.0 \,\mathrm{kg\,m^{-3}}$, ``white''), a situation that persists further down the shelf (\fref{fig:CrossSectionsSpice}c, d).  The onshore waters ($-5 < X < 40\, \mathrm{km}$) are remarkably well-mixed in $\theta$--$S$ space, despite still retaining significant density stratification.

@@

Oxygen profiles, computed in the same manner, show similar trends, with the low-spice anomaly water clearly having a very low oxygen signature, while the strong negative spice anomaly shallower than $\sigma_{\theta} = 26.4\,\mathrm{kg\,m^{-3}}$ has very high oxygen, and the deeper water is also more oxygenated.  Of note is that compared with the coarser sampling along the LC Line (\fref{fig:LaPerouse2013Ctd}), we see that there is a very abrupt transition between the low spice anomaly/low-O2 water and the offshore water over the ``Eddy'' region.  We will see below that this transition is even sharper than indicated by these synthetic sections, since they alias some lateral and temporal irregularity in the $\theta$--$S$-compensated front.

@@

The spatial patterns are very clear when considering a map-view of properties along the $\sigma_{\theta} = 26.4\,\mathrm{kg\,m^{-3}}$ isopycnal (\fref{fig:SpiceO2264}).  There is a region onshore of the south tip of La Perouse Bank (approximately 44.5 N, 125.75 W), that consists of water that is found along the mixing line (spice anomaly $\gamma \approx 0 \mathrm{kg\,m^{-3}}$) with low oxygen saturations; this water is called the Juan de Fuca eddy (or Tully eddy), though here an eddy shape is not readily apparent.  Offshore of this region is water that is very warm and salty in comparison ($\gamma > 0.15 \mathrm{kg\,m^{-3}}$), and relatively high in oxygen (saturations of approximately 60\%).  The region between these two water masses is very abrupt, and is concave with respect to the isobaths on the shelf,  stretching from La Perouse Bank to a shallow bank just above the Juan de Fuca canyon (48.3 N and 125.4 W).

Upstream of La Perouse bank, and Barkley Canyon, the water along the shelf has a weaker spice anomaly than the water further downstream ($\gamma \approx 0.7$, light pink colors).  This water is also somewhat lower in oxygen than water from offshore, though not as depleted as the water in the "eddy".  This water appears to be pushed offshore just upstream of Barkley Canyon, and replaced on the shelf by the warmer offshore water.

@@

\subsubsection{Frontal survey}

A systematic survey through the front between the offshore water and the ``eddy'' water demonstrates the structure of the front (\fref{fig:Frontsurvey}).  The survey started close to shore and passed through the Vancouver Island Coastal Current (casts 2030--2080).  This water is buoyant and signficiantly fresher and cooler than water at the same density near the surface.  There is strong evidence of isopycnal mixing of the surface water and the outflow water, with warm-salty water in the relatively wide front between the ``eddy'' water and the water in the current.

>>>>

\title{Separation of an upwelling current bounding the Juan de Fuca Eddy}

@@

\authors{Jody M. Klymak\affil{1,2}, Susan E. Allen\affil{3,4}, Stephanie N. Waterman\affil{3}}

\affiliation{1}{School of Earth and Ocean Sciences, University of Victoria, BC, Canada}
\affiliation{2}{Department of Physcics \& Astronomy, University of Victoria, BC, Canada}
\affiliation{3}{Department of Earth, Ocean and Atmospheric Sciences, University of British Columbia, Vancouver, BC, Canada}
\affiliation{4}{Institute of Applied Mathematics, University of British Columbia, Vancouver, BC, Canada}
%(repeat as many times as is necessary)

@@

\begin{keypoints}
\item The shelf break current along Vancouver Island separates downstream of a submarine bank.
\item Offshore water is drawn onto the shelf and forms a sharp semi-persistent front with the Juan de Fuca Eddy.
\item The Eddy shows evidence of long residence times, and little evidence of deep-water origin.
\end{keypoints}

@@

Comprehensive water column observations of temperature, salinity, and oxygen on the productive southern Vancouver Island shelf offer a unique characterization of processes affecting exchange on the shelf.  A semi-permanent cyclonic recirculation (the Juan de Fuca Eddy) occupies the region in the lee of a bank where the shelf widens abruptly.  The observations indicate that the water in this Eddy is a mixture of offshore water and water from a buoyant coastal current.  The water in the  in the Eddy is well-mixed along a mixing line in $\theta$--$S$ space, though it retains stratification, and is either rapidly mixed (high $\kappa$), or has a long residence time (large $\tau$), such that $(\kappa \tau)^{1/2} \sim 50 -- 100\ \mathrm{m}$.  There is a sharp temperature-salinity front on the offshore side of of this well-mixed water, with the front no more than 1-km wide, and no sign of instabilities. Previous work had hypothesized the role of a spur canyon incising the shelf in feeding water to the Eddy; little evidence of this was found during the observations, however, these were collected during a relaxation and reversal of upwelling winds, which may have slowed the flow of water into the Eddy.  Tidally resolved observations in the spur canyon also show very strong hydraulic flows in the cross-canyon direction accompanied by very high localized mixing rates in 50-m hydraulic jumps possibly exceeding $\kappa \sim 1 \ \mathrm{m^2s^{-1}}$.

Upstream of the Eddy, there is an along-shelf current flowing equatorward with partially mixed water properties compared to the well-mixed water in the Eddy.  The shelf current was observed to separate from the shelf in the lee of a bank, and is replaced by the distinct offshore water mass that forms the offshore front with the Eddy.  The separation event can be seen in sea-surface temperatures in satellite images as tongue of water that is ejected offshore, possibly with the relaxation of upwelling, with the flow heading directly offshore.  The cause of this offshore separation event is not determined, but likely involves instability between the shelfbreak current and the California Undercurrent, and possibly is catalyzed by separation from the abrupt underwater bank at the poleward end of the study region.

@@

Cross-shelf exchange is important to the health and productivity of continental shelf regions, where offshore oxygenated water is exchanged with nutrient-rich, but oxygen-depleted nearshore water.  Cross-shelf transport usually requires ageostrophic flow, since geostrophically balanced flow will tend to follow topographic contours, often providing a barrier to lateral exchange \cite{brink16}.  Mechanisms for cross-shelf exchange include internal waves and instabilities in shelf-break fronts.  Most-dramatically, however, topography can drive cross-slope exchanges, often seen most clearly in surface imagery, but occasionally observed \emph{in-situ} \cite{barthetal00}.

Here we present detailed \emph{in-situ} observations from the southern Vancouver Island shelf, collected in summer 2013, and sampled with a rapid profiling vehicle equipped with a CTD and oxygen sensor, supplemented by traditional hydrographic surveys. The shelf has complicated bathymetry \fref{fig:LocMapBoth}, with a somewhat typical shelf poleward of our study site, but then widening in the lee of La Perouse Bank, and then hitting the Juan de Fuca canyon, equatorward.  Water from the strait of Juan de Fuca flows poleward as a buoyant current that hugs the coast \cite{thomsonetal89, hickeyetal91}, while shelf water flows equatorward in the summer under upwelling conditions, both forced by local winds and via tele-connections with the long, homogenous shelves equatorward off Washington and Oregon \cite{hickeyetal91,thomsonkrassovski15,engidaetal16}.  Trapped between these two currents is a region of relatively homogenous water that has been termed the Juan de Fuca Eddy \cite{freelanddenman82,freelandmcintosh89,foremanetal07,macfadyenhickey10}.

A goal of our study was to understand how the Juan de Fuca Eddy persists, and how it exchanges water properties with offshore water.  Low oxygen has previously been found in the Eddy, causing it to be inferred that water in the eddy is upwelled from as deep as 400 m \cite{freelanddenman82,deweycrawford88}.  This inference was based on the oxygen being conservative, and hence having to come from close to the oxygen minimum zone found further offshore \cite{mackasetal87} and lead to hypotheses that perhaps the eddy was fed by waters being drawn up a spur canyon (called the Spur Canyon) via ageostrophic transport from offshore to onshore due to the low pressure in the eddy center \cite{weaverhsieh87}.  Below, we will argue that there is not substantial evidence of such transport, and that oxygen is likely low because of consumption due to respiration.

A second goal was to better understand cross-shore exchange with offshore water.  While the typical schematic of shelf exchange is that there are Ekman layers near the boundaries, there is also evidence of filaments and wholesale separation of shelf currents into the deep ocean, both at this site, and at others.  Such events are important as they tend to transport nutrient rich shelf water out into the open ocean, fueling offshore ecosystems.  Along the Vancouver Island shelf, large filaments or instabilities of the coastal current have been inferred using satellite images \cite{ikedaemery84,thomsongower98}.  Direct observations of such filaments have been made further to the south off Cape Blanco \cite{barthetal00}, and other coasts \cite[e.g.]{relvasbarton05}. Net cross-shore transport of nutrients and chlorophyll have been found in hydrographic surveys off Vancouver Island \cite{mackasyelland99}, and associated with mesoscale features both in geostrophic velocities, and in satellite observations.  The mechanisms driving such separations are poorly understood, but hypotheses include coastal hydraulics \cite{dalebarth01} leading to along-shore trapping of coastal waves,  wind stress curl variations \cite{castelaobarth07}, induced relative vorticity due to stretching of parcels that inertially overshoot their initial isobaths due to a sudden change in downstream bathymetry, and the non-linear breaking of large-scale meanders due to baroclinic instability between the wind-driven current and the California Undercurrent \cite{ikedaetal84, batteen97}.

In this paper we present  observations from 2013 (\fref{sec:Site}), presenting the data in a number of ways to highlight the important processes, with a focus on the offshore side of the eddy region (\fref{sec:Observations}) where we consider the age of the eddy, the steadiness of the front separating the eddy and the offshore water, properties along the spur canyon that was believed to feed the eddy, and finally describe a large offshore tongue of the shelf-break current and its intrusion into the interior.  We discuss the origin of the eddy and the implications of the separating jet (\fref{sec:Summary}).

@@

We consider routine hydrographic surveys along the LB and LC lines, collected aboard the \emph{CCGS Tully} from 2013-05-30 to 2013-05-31  (``June''), and 2013-09-07 to 2013-09-09 (``September'').  These lines span the depths from about 50 m to deep offshore, with casts every 7.5 km across shelf. The data comes from a lowered Seabird 9-11 CTD, with an SBE 43 oxygen sensor.  Oxygen data have been checked against bottle casts, and the CTD corrected for sensor offsets and thermal lags.

We focus on fine-scale surveys carried out from 2013-08-21 to 2013-08-30.  Data were collected from the \emph{R/V Falkor} with an ODIM Brooke-Ocean Moving Vessel Profiler (MVP).  The MVP was equipped with an AML Oceanographic CTD, and a Rinko Oxygen sensor with a 7-s response time foil.  The MVP profiles to depths of 200 m or to within 5 m of the seafloor, whichever is shallower, and drops at a speed of $3\ \mathrm{m\,s^{-1}}$.  Data collection took place while the ship cruised at speeds between 5 and 8 kts, usually at around 6 kts to enable fine horizontal spacing of the casts, with typical spacing less than 800 m.  Data is reported for the downcast, which mostly follows a vertical path.  The rapid speed of the profiling makes the oxygen measurements somewhat coarse, and probably biased due to the phase lag of the sensor, so we treat these largely qualitatively in this paper.

@@

Winds during the cruises were typical for the west coast of Vancouver Island, with upwelling favorable winds during July and early August (\fref{fig:LaPeWind}).  During the finescale cruise, and for the week previous, the winds were intermittently downwelling favorable.  Note that this locale is strongly affected by coastally trapped waves from further south, so doming of near-bottom isopycnals often persists despite local wind forcing \cite{thomsonkrassovski15, engidaetal16}, and takes a finite amount of time to spin down.

\begin{figure*}[htbp]
  \begin{center}
    \includegraphics[width=5.5in]{LaPeWind}
    \caption{
      Wind from the La Perouse buoy \cite{DFOWind2013C46206}.  The vertical direction is along-shelf (chosen as a heading of 315 N), so vectors pointing straight down represent upwelling winds.  .  The wind components has been low-pass filtered to one-day averages.  The timing of the cruises discussed in the paper are shown as colored bands.
      \label{fig:LaPeWind} }
  \end{center}
\end{figure*}

@@

Hydrographic sections along the LB and LC lines highlight summer conditions on the Southern Vancouver Island Shelf, and indicate some of the features we are focusing on in this paper (\Fref{fig:LaPerouse2013Ctd}).  During both cruises and at both lines there is clear evidence of upwelling, with the $26.4\ \mathrm{kg\,m^{-3}}$ isopycnal reaching from 130 m depth to shallower than 85 m over the shelf.  This dense water tends to be low in oxygen and cool. Further onshore, isopycnals tilt back down, and the water warms, evidence of the Vancouver Island Coastal Current that hugs the coast here.

\begin{figure*}[htbp]
  \begin{center}
     \includegraphics[width=130mm]{LaPerouse2013Ctd}
     \caption{Observations along LC (top three rows) and LB lines (bottom three rows) in Jun (left column) and Sep 2013 (right column).  X is across-shelf distance in the coordinate system shown in \fref{fig:LocMapBoth}. Potential density is contoured every $0.2\ \mathrm{kg\,m^{-3}}$, with the $26.4\ \mathrm{kg\,m^{-3}}$ colored magenta.  Potential temperature, $\theta$, spice anomaly, $\gamma$ and oxygen concentration, $O_2$, are colored for each section.}
     \label{fig:LaPerouse2013Ctd}
  \end{center}
\end{figure*}

The Juan de Fuca Eddy region is along the LB Line (\fref{fig:LaPerouse2013Ctd}, bottom three rows), mostly inshore of 0 km. This water is cooler and lower in oxygen than along the same isopycnal further offshore.  Also note that during the June cruise the water in this region was cooler and more oxygenated than later in the summer during the September cruise.  Near the surface, the isopycnals are domed, but the low-oxygen anomaly extends as high as the thin surface mixed layer.

The water upstream of the Eddy region, along the LC line appears less mixed than the Eddy water.  It does not show as much drawdown of oxygen, and is warmer, at least offshore of La Perouse bank (\fref{fig:LaPerouse2013Ctd}, top three rows).  Water does not appear to make it over La Perouse bank into the basin onshore, or if it does, it does so intermittently.  There appears to be some mixing of the deeper and shallow water near $26.4\ \mathrm{kg\,m^{-3}}$, with the water along this isopycnal becoming cooler and lower in oxygen where it intersects the shelf, consistent with enhanced mixing near the shelf

The $\theta$--$S$ properties show many of the features demonstrated above (\fref{fig:LaPerouse2013TS}a).  Below $26.6\ \mathrm{kg\,m^{-3}}$ or so, the deep water masses are found offshore, and are largely homogenous. Shallower, there are distinct differences between the water found offshore, with it being warmer and saltier than water on the shelf at the same densities. These anomalies are hard to parse directly, so in this paper we use a spice anomaly, relative to a straight line in $\theta$--$S$ space (\fref{fig:LaPerouse2013TS}). This line represents an approximate mixing line for water in the Eddy, and passes the points $7.75\ \mathrm{^oC}$, $30\ \mathrm{psu}$ and $6.6\ \mathrm{^oC}$, $35\ \mathrm{psu}$. Spice anomaly for a given water sample at density $\sigma_{\theta}$ is given by $\gamma = \alpha \left(T - T_0\left(\sigma_{\theta}\right)\right) + \beta \left(S - S_0\left(\sigma_{\theta}\right)\right)$ where $T_0\left(\sigma_{\theta}\right)$, where $S_0\left(\sigma_{\theta}\right)$ are the temperature and salinity along the mixing line at the same density as the water sample, with the sign convention that positive is warmer and saltier than data along the mixing line.

\begin{figure}[htbp]
  \begin{center}
     \includegraphics[width=5in]{LaPerouse2013TS}
     \caption{Potential temperature versus salinity for the La Perouse data shown in \fref{fig:LaPerouse2013Ctd}; potential density is contoured every $0.2\ \mathrm{kg\,m^{-3}}$, with the $26.4\ \mathrm{kg\,m^{-3}}$ colored magenta. A definition of spice anomaly is shown in this plot, and discussed in the text.  The rectangle is the $\theta-S$ range used in \fref{fig:TSdensSpice} below.}
     \label{fig:LaPerouse2013TS}
  \end{center}
\end{figure}

Offshore water tends to be high is absolute spice anomaly (\fref{fig:LaPerouse2013Ctd}, middle rows), with a positive-anomaly layer (red) centered at $26.4\ \mathrm{kg\,m^{-3}}$ sandwiched between negative-anomaly layers above and below.  On the shelf ($-10\, \mathrm{km} < X < 30\, \mathrm{km}$), the distinct $\theta$--$S$ masses are attenuated and much closer to the defined mixing line than the offshore water (closer to white in color).  Hugging the coast ($X > 30\, \mathrm{km}$), is the Vancouver Island Coastal Current, with a negative spice anomaly in June, and a positive anomaly in summer, consistent with seasonal warming.

@@

     \includegraphics[width=3.5in]{TSdensSpice.pdf}
    \caption{Sample density of salinity and potential density data from the cruise, with a logarithmic color scale.  Grey contours are potential density relative to the surface, every $0.1\ \mathrm{kg\,m^{-3}}$, and the magenta contour is the $26.4\ \mathrm{kg\,m^{-3}}$ isopycnal.  Spice anomaly, $\gamma$, is defined as temperature anomaly scaled by the thermal expansion co-efficient, $\alpha$, along isopycnals from the mixing line found in the well-mixed region in the Eddy (white line), in $\mathrm{kg\,m^{-3}}$.
    \label{fig:TSdensSpice}}
  \end{center}
\end{figure*}

In synthetic cross sections of spice anomaly, the Eddy is clearly identifiable as water that is along the straight mixing line, and thus has low-spice anomaly ($\gamma \approx 0\ \mathrm{kg\,m^{-3}}$, \fref{fig:CrossSectionsSpice}c, d).  This signature of well-mixed water extends from the seafloor to approximately 30 m depth, onshore of $x \approx 0\ \mathrm{km}$.  Despite lying along a mixing line, the Eddy water is still stratified.  This well-mixed water contrasts with water directly offshore on the other side of a sharp $\theta-S$ compensated front that is warmer and saltier than the well mixed water ($\gamma > 0.2\ \mathrm{kg\,m^{-3}}$) deeper than $26\,\mathrm{kg\,m^{-3}}$, and colder and fresher shallower.  The thickness of this front is even sharper in individual sections than in these composite sections (see \fref{sec:frontsurvey}).

In contrast, the third population of partially mixed water mass identified in \fref{fig:TSdensSpice} is all found upstream of the Eddy region (\fref{fig:CrossSectionsSpice}a, b) as a weaker spice anomaly ($\gamma \approx 0.1\ \mathrm{kg\,m^{-3}}$, ``pink'' colors).  This water is generally found near the intersection of the upwelling water and La Perouse Bank, and is not as warm as offshore water, indicating some mixing has taken place.  Of note, is that this intermediate water mass appears completely absent in the sections further equatorward (\fref{fig:CrossSectionsSpice}c, d), indicating that it is either mixed away, or that it is advected elsewhere; we argue below that it is advected offshore in a large-scale tongue.

@@

Oxygen section show similar trends, with the low-spice anomaly water having low oxygen saturations, while the strong negative spice anomaly shallower than $\sigma_{\theta} = 26.4\,\mathrm{kg\,m^{-3}}$ has very high oxygen, and the deeper water is also more oxygenated.  Of note is that compared with the coarser sampling along the LC Line (\fref{fig:LaPerouse2013Ctd}), we see that the abrupt transition between offshore and Eddy water is also present in oxygen.

@@

The spatial patterns are very clear when considering a map-view of properties along the $\sigma_{\theta} = 26.4\,\mathrm{kg\,m^{-3}}$ isopycnal (\fref{fig:SpiceO2264}).  There is a region onshore of the south tip of La Perouse Bank (approximately 44.5 N, 125.75 W), that consists of water that is found along the mixing line (spice anomaly $\gamma \approx 0 \mathrm{kg\,m^{-3}}$) with low oxygen saturations.  Offshore of this region is water that is very warm and salty in comparison ($\gamma > 0.15 \mathrm{kg\,m^{-3}}$), and relatively high in oxygen (saturations of approximately 60\%).  The region between these two water masses is very abrupt, and is concave with respect to the isobaths on the shelf,  stretching from La Perouse Bank to a shallow bank just above the Juan de Fuca canyon (48.3 N and 125.4 W).

Upstream of La Perouse bank and Barkley Canyon, the water along the shelf the weaker spice anomaly of the third water mass ($\gamma \approx 0.1\,\mathrm{kg\,m^{-3}}$, light pink colors).  This water is also somewhat lower in oxygen than water from offshore, though not as depleted as the water in the Eddy.  This water appears to be pushed offshore just upstream of Barkley Canyon, and replaced on the shelf by the warmer offshore water.

@@

\subsubsection{Frontal survey}
\label{sec:frontsurvey}

A systematic survey through the front between the offshore water and the Eddy water demonstrates the structure of the front (\fref{fig:Frontsurvey}).  The survey started close to shore and passed through the Vancouver Island Coastal Current (along-track 0-10 km).  The coastal current water is buoyant and significantly fresher and cooler than water at the same density near the surface and forms a classic density front with

The front between the offshore water and the ``eddy'' water was passed through 10 times, showing its evolution following the along-shore equatorward flow.  First, as noted in the composite sections, isopycnals slope up from offshore onto the shelf.  This is in agreement with a shelf-break current, though the isopycnals flatten out in the Eddy region.  The front starts out very sharp, (along-track distance $50\ \mathrm{km}$) though two small tendrils can be seen separating from the front on the inshore side.  Similar tendrils are found on the second pass, perhaps a bit more separated from the front ($\approx 80\ \mathrm{km}$), and on the third pass ($\approx 95\ \mathrm{km}$).  Note that these tendrils are made of up partially mixed water.  The subsequent passes have more of this partially mixed water, such that the partially mixed region is up to 5-km wide by the fifth pass ($\approx 150\ \mathrm{km}$).  However, note that the deeper isopycnals retain a sharp front, and indeed the front appears sharp again by the seventh pass at all depths.
